---
title: "AlphaFold Analysis"
author: "Yaniv Iny (PID: A18090586)"
format: pdf
---

Here we analyze our AlphaFold structure prediction models. The input directory/folder comes from the ColabFold server: 

```{r}
# Change this for YOUR results dir name
results_dir <- "CORRECTHIVMODEL_94b5b"
```

```{r}
# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

I will use the Bio3D package for analysis & Align and superpose
```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
pdbs
```
A quick PCA plot 

```{r}
pc <- pca(pdbs)
plot(pc)
```
RMSD Analysis 

RMSD is a common measure of structural distance used in structural biology
```{r}
rd <- rmsd(pdbs, fit=T)
```
```{r}
rd
```
'
```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")


```
```{r}
core <- core.find(pdbs)
```
```{r}
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

Predicted Alignment error for domains 

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```
```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

```{r}
pae1$max_pae

```

```{r}
pae5$max_pae

```
```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```
```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```
Reisudue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
dim(aln$ali)
```

```{r}
sim <- conserv(aln)
```
```{r}
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```
```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```


