---
title: "Class 10 Comparative Structural Analysis"
author: "Yaniv Iny (PID:A18090586)"
format: html
---

## Comparative structure analysis of Adenylate Kinase

Here we run through a complete "pipeline" of strucutre analysis that begins with a single sequence identifierf and ends in a PCA analysis. 

```{r}
library(bio3d)
```

## Step 1. retrieve a sequence for the protien we are interested in. We willl take ADK "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```
 ## STEP 2 Blast search 
Run a BLAST search of the PDB for all related sequences to our input `aa`

```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```
## Step 3 Download all Structures
These are our "top hits" i.e. all the structures in the pDB databe related to our input sequence. 

```{r}
hits$pdb.id
```

```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

## Step 4 align and superimpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Step 5. PCA 

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 

```{r}
pc <- pca(pdbs)
plot(pc)
```

lets make a trajectory (or movie)of the main conformation changes captures by PC1. We will use the `mktrj()` function fot this... 

```{r}
mktrj(pc, file="pca_results.pdb")
```

Back of the envelope comparison of PDB size to Uniprot

```{r}
uniprot <- 253206171
pdb <- 231029
231029/253206171 * 100
```

